---
title: "Mental Health & Suicide Dashboard (2016 Data)"
author: "Fazi_DA"
output:
  flexdashboard::flex_dashboard:
    source_code: embed
---

```{r setup, include = FALSE}
#libraries
library(mice)
library(tidyverse)
library(reshape2)
library(plotly)
library(flexdashboard)

#read data
age_rate <- read.csv('Age-standardized-suicide-rates.csv')
crude_rate <- read.csv('Crude-suicide-rates.csv')
facilities <- read.csv('Facilities.csv')
HR <- read.csv('Human-Resources.csv')
happy <- read.csv('whr-2016.csv')
#all data for year 2016
#first four is data per 100,000 population

####Processing####

#facilities
facilities <- subset(facilities, select = -c(Year))
facilities <- facilities[rowSums(is.na(facilities[,1:6])) <= 2, ]
facilities_temp <- mice(data = facilities, m = 10, method = "pmm", maxit = 50,
                        seed = 1234)
facilities_complete <- complete(facilities_temp, 1)

#HR
HR <- subset(HR, select = -c(Year))
HR <- HR[rowSums(is.na(HR[,1:5])) <= 2, ]
HR_temp <- mice(data = HR, m = 10, method = "pmm", maxit = 50, seed = 1234)
HR_complete <- complete(HR_temp, 1)

#age_rate
names(age_rate)[names(age_rate) == 'X2016'] <- '2016'
age_rate <- age_rate[, 1:3]
age_rate <- reshape(age_rate, idvar = 'Country', timevar = 'Sex', direction = 'wide')
colnames(age_rate)[2] <- 'Male&Female'
colnames(age_rate)[3] <- 'Male'
colnames(age_rate)[4] <- 'Female'

#happy
happy <- happy[,c(-3, -5,-6)]

#combine data
all_data <- happy %>%
  inner_join(facilities_complete, by = 'Country') %>%
  inner_join(HR_complete, by = 'Country') %>%
  inner_join(age_rate, by = 'Country')
colnames(all_data)[20] <- 'Suicide_Rate_MF'
colnames(all_data)[4] <- 'Economy'
colnames(all_data)[6] <- 'Health'
colnames(all_data)[8] <- 'Gov.Corruption'
all_data <- all_data[,c(-2,-21,-22)]

#crude_rate
for (i in seq(1:10)){
 colnames(crude_rate)[i] <- gsub('X', '', colnames(crude_rate)[i]) 
}
crude_rate <- crude_rate %>%
  filter(Country %in% c(all_data$Country))
crude_rate <- melt(crude_rate, id.vars = c('Country', 'Sex'), value.name = 'Rate')
colnames(crude_rate)[3] <- 'Age_Band'
crude_rate$Country <- reorder(crude_rate$Country, crude_rate$Rate)
```


Age Groups
=======================================================================

Column {data-width = 650}
-----------------------------------------------------------------------

### Suicide Rates by Age Group (per gender) (per 100,000 population)

```{r warning=FALSE}
ggplotly(ggplot(subset(crude_rate, Sex == ' Male'|Sex == ' Female'), 
       aes(x = Country, y = Rate)) +
  geom_bar(aes(fill = Age_Band), position = 'stack', stat = 'identity') +
  theme_minimal() +
  xlab(" ") +
  ylab(" ") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_text(size = 7)) +
  scale_fill_brewer(palette = 'RdBu') +
  facet_wrap(~Sex) +
  coord_flip())
```


Column {data-width = 350}
-----------------------------------------------------------------------

### Suicide Rates by Age Group (both genders) (per 100,000 population)

```{r warning = FALSE}
ggplotly(ggplot(subset(crude_rate, Sex == ' Both sexes'), 
       aes(x = Country, y = Rate)) +
  geom_bar(aes(fill = Age_Band), stat = 'identity', position = 'stack') +
  theme_minimal() +
  xlab(" ") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_text(size = 7)) +
  scale_fill_brewer(palette = 'RdBu') +
  coord_flip())
```

Happiness vs Suicide
=======================================================================

### Happiness Score vs Suicide Rates (per 100,000 population)

```{r warning=FALSE}
ggplotly(ggplot(all_data, aes(x = Happiness.Score, y = Suicide_Rate_MF, Country = Country)) +
  geom_point(size = 3, 
             color = case_when(all_data$Happiness.Score >= 6 ~ 'blue',
                               all_data$Suicide_Rate_MF >= 15 ~ 'red',
                               TRUE ~ 'black')) +
  ylab("Suicide Rate (Male & Female)") +
  xlab("Happiness Score") +  
  theme_minimal())
```

Country Data
=====================================================================

Column {.tabset}
---------------------------------------------------------------------

### Mental Hospitals

```{r warning=FALSE}
all_data$Country <- reorder(all_data$Country, all_data$Mental._hospitals)
ggplotly(ggplot(all_data, aes(x = Country, y = Mental._hospitals, 
                              Suicide_Rate = Suicide_Rate_MF, Happines_Score = Happiness.Score)) +
           geom_bar(stat = 'identity', color = 'black', fill = 'firebrick1') +
           xlab("Country") +
           ylab("Mental Hospitals per 100,000") +
           theme_minimal() +
           theme(axis.text.x = element_blank(),
                 axis.text.y = element_text(size = 7)) +
           coord_flip()) 
```

### Health Units

```{r warning=FALSE}
all_data$Country <- reorder(all_data$Country, all_data$health_units)
ggplotly(ggplot(all_data, aes(x = Country, y = health_units, 
                              Suicide_Rate = Suicide_Rate_MF, Happines_Score = Happiness.Score)) +
           geom_bar(stat = 'identity', color = 'black', fill = 'firebrick1') +
           xlab("Country") +
           ylab("Health Units per 100,000") +
           theme_minimal() +
           theme(axis.text.x = element_blank(),
                 axis.text.y = element_text(size = 7)) +
           coord_flip()) 
```

### Outpatient Facilities

```{r warning=FALSE}
all_data$Country <- reorder(all_data$Country, all_data$outpatient._facilities)
ggplotly(ggplot(all_data, aes(x = Country, y = outpatient._facilities, 
                              Suicide_Rate = Suicide_Rate_MF, Happines_Score = Happiness.Score)) +
           geom_bar(stat = 'identity', color = 'black', fill = 'firebrick1') +
           xlab("Country") +
           ylab("Outpatient Facilities per 100,000") +
           theme_minimal() +
           theme(axis.text.x = element_blank(),
                 axis.text.y = element_text(size = 7)) +
           coord_flip()) 
```

### Psychiatrists

```{r warning=FALSE}
all_data$Country <- reorder(all_data$Country, all_data$Psychiatrists)
ggplotly(ggplot(all_data, aes(x = Country, y = Psychiatrists, 
                              Suicide_Rate = Suicide_Rate_MF, Happines_Score = Happiness.Score)) +
           geom_bar(stat = 'identity', color = 'black', fill = 'firebrick1') +
           xlab("Country") +
           ylab("Psychiatrists per 100,000") +
           theme_minimal() +
           theme(axis.text.x = element_blank(),
                 axis.text.y = element_text(size = 7)) +
           coord_flip()) 
```

### Nurses

```{r warning=FALSE}
all_data$Country <- reorder(all_data$Country, all_data$Nurses)
ggplotly(ggplot(all_data, aes(x = Country, y = Nurses, 
                              Suicide_Rate = Suicide_Rate_MF, Happines_Score = Happiness.Score)) +
           geom_bar(stat = 'identity', color = 'black', fill = 'firebrick1') +
           xlab("Country") +
           ylab("Nurses per 100,000") +
           theme_minimal() +
           theme(axis.text.x = element_blank(),
                 axis.text.y = element_text(size = 7)) +
           coord_flip()) 
```

### Social Workers

```{r warning=FALSE}
all_data$Country <- reorder(all_data$Country, all_data$Social_workers)
ggplotly(ggplot(all_data, aes(x = Country, y = Social_workers, 
                              Suicide_Rate = Suicide_Rate_MF, Happines_Score = Happiness.Score)) +
           geom_bar(stat = 'identity', color = 'black', fill = 'firebrick1') +
           xlab("Country") +
           ylab("Social Workers per 100,000") +
           theme_minimal() +
           theme(axis.text.x = element_blank(),
                 axis.text.y = element_text(size = 7)) +
           coord_flip()) 
```

Column {.tabset}
---------------------------------------------------------------------

### Economy Score

```{r warning=FALSE}
all_data$Country <- reorder(all_data$Country, all_data$Economy)
ggplotly(ggplot(all_data, aes(x = Country, y = Economy, 
                              Suicide_Rate = Suicide_Rate_MF, Happines_Score = Happiness.Score)) +
           geom_bar(stat = 'identity', color = 'black', fill = 'royalblue') +
           xlab("Country") +
           ylab("Economy Score") +
           theme_minimal() +
           theme(axis.text.x = element_blank(),
                 axis.text.y = element_text(size = 7)) +
           coord_flip()) 
```

### Family Score

```{r warning=FALSE}
all_data$Country <- reorder(all_data$Country, all_data$Family)
ggplotly(ggplot(all_data, aes(x = Country, y = Family, 
                              Suicide_Rate = Suicide_Rate_MF, Happines_Score = Happiness.Score)) +
           geom_bar(stat = 'identity', color = 'black', fill = 'royalblue') +
           xlab("Country") +
           ylab("Family Score") +
           theme_minimal() +
           theme(axis.text.x = element_blank(),
                 axis.text.y = element_text(size = 7)) +
           coord_flip()) 
```

### Health Score

```{r warning=FALSE}
all_data$Country <- reorder(all_data$Country, all_data$Health)
ggplotly(ggplot(all_data, aes(x = Country, y = Health, 
                              Suicide_Rate = Suicide_Rate_MF, Happines_Score = Happiness.Score)) +
           geom_bar(stat = 'identity', color = 'black', fill = 'royalblue') +
           xlab("Country") +
           ylab("Health Score") +
           theme_minimal() +
           theme(axis.text.x = element_blank(),
                 axis.text.y = element_text(size = 7)) +
           coord_flip()) 
```

### Freedom Score

```{r warning=FALSE}
all_data$Country <- reorder(all_data$Country, all_data$Freedom)
ggplotly(ggplot(all_data, aes(x = Country, y = Freedom, 
                              Suicide_Rate = Suicide_Rate_MF, Happines_Score = Happiness.Score)) +
           geom_bar(stat = 'identity', color = 'black', fill = 'royalblue') +
           xlab("Country") +
           ylab("Freedom Score") +
           theme_minimal() +
           theme(axis.text.x = element_blank(),
                 axis.text.y = element_text(size = 7)) +
           coord_flip()) 
```

### Government Corruption Score

```{r warning=FALSE}
all_data$Country <- reorder(all_data$Country, all_data$Gov.Corruption)
ggplotly(ggplot(all_data, aes(x = Country, y = Gov.Corruption, 
                              Suicide_Rate = Suicide_Rate_MF, Happines_Score = Happiness.Score)) +
           geom_bar(stat = 'identity', color = 'black', fill = 'royalblue') +
           xlab("Country") +
           ylab("Government Corruption Score") +
           theme_minimal() +
           theme(axis.text.x = element_blank(),
                 axis.text.y = element_text(size = 7)) +
           coord_flip()) 
```

### Generosity Score

```{r warning=FALSE}
all_data$Country <- reorder(all_data$Country, all_data$Generosity)
ggplotly(ggplot(all_data, aes(x = Country, y = Generosity, 
                              Suicide_Rate = Suicide_Rate_MF, Happines_Score = Happiness.Score)) +
           geom_bar(stat = 'identity', color = 'black', fill = 'royalblue') +
           xlab("Country") +
           ylab("Generosity Score") +
           theme_minimal() +
           theme(axis.text.x = element_blank(),
                 axis.text.y = element_text(size = 7)) +
           coord_flip()) 
```

Facilities vs Suicide Rate
=====================================================================

Row
---------------------------------------------------------------------

### Mental Hospitals vs Suicide Rates (per 100,000 population)

```{r warning=FALSE}
ggplotly(ggplot(all_data, aes(x = Mental._hospitals, y = Suicide_Rate_MF, Country = Country)) +
  geom_point(size = 2) +
  ylab("Suicide Rate (Male & Female)") +
  xlab("Mental Hospitals") +  
  theme_minimal())
```

### Health Units vs Suicide Rates (per 100,000 population)

```{r warning=FALSE}
ggplotly(ggplot(all_data, aes(x = health_units, y = Suicide_Rate_MF, Country = Country)) +
  geom_point(size = 2) +
  ylab("Suicide Rate (Male & Female)") +
  xlab("Health Units") +  
  theme_minimal())
```

### Outpatient Facilities vs Suicide Rates (per 100,000 population)

```{r warning=FALSE}
ggplotly(ggplot(all_data, aes(x = outpatient._facilities, y = Suicide_Rate_MF, Country = Country)) +
  geom_point(size = 2) +
  ylab("Suicide Rate (Male & Female)") +
  xlab("Outpatient Facilities") +  
  theme_minimal())
```

Row
--------------------------------------------------------------------

### Psychiatrists vs Suicide Rates (per 100,000 population)

```{r warning=FALSE}
ggplotly(ggplot(all_data, aes(x = Psychiatrists, y = Suicide_Rate_MF, Country = Country)) +
  geom_point(size = 2) +
  ylab("Suicide Rate (Male & Female)") +
  xlab("Psychiatrists") +  
  theme_minimal())
```

### Nurses vs Suicide Rates (per 100,000 population)

```{r warning=FALSE}
ggplotly(ggplot(all_data, aes(x = Nurses, y = Suicide_Rate_MF, Country = Country)) +
  geom_point(size = 2) +
  ylab("Suicide Rate (Male & Female)") +
  xlab("Nurses") +  
  theme_minimal())
```

### Social Workers vs Suicide Rates (per 100,000 population)

```{r warning=FALSE}
ggplotly(ggplot(all_data, aes(x = Social_workers, y = Suicide_Rate_MF, Country = Country)) +
  geom_point(size = 2) +
  ylab("Suicide Rate (Male & Female)") +
  xlab("Social Workers") +  
  theme_minimal())
```


Happiness Score Contributors vs Suicide Rate
=====================================================================

Row
---------------------------------------------------------------------

### Economy Score vs Suicide Rates

```{r warning=FALSE}
ggplotly(ggplot(all_data, aes(x = Economy, y = Suicide_Rate_MF, Country = Country)) +
  geom_point(size = 2) +
  ylab("Suicide Rate (Male & Female)") +
  xlab("Economy") +  
  theme_minimal())
```

### Family Score vs Suicide Rates

```{r warning=FALSE}
ggplotly(ggplot(all_data, aes(x = Family, y = Suicide_Rate_MF, Country = Country)) +
  geom_point(size = 2) +
  ylab("Suicide Rate (Male & Female)") +
  xlab("Family") +  
  theme_minimal())
```

### Health Score vs Suicide Rates 

```{r warning=FALSE}
ggplotly(ggplot(all_data, aes(x = Health, y = Suicide_Rate_MF, Country = Country)) +
  geom_point(size = 2) +
  ylab("Suicide Rate (Male & Female)") +
  xlab("Health") +  
  theme_minimal())
```

Row
--------------------------------------------------------------------

### Freedom Score vs Suicide Rates

```{r warning=FALSE}
ggplotly(ggplot(all_data, aes(x = Freedom, y = Suicide_Rate_MF, Country = Country)) +
  geom_point(size = 2) +
  ylab("Suicide Rate (Male & Female)") +
  xlab("Freedom") +  
  theme_minimal())
```

### Government Corruption Score vs Suicide Rates 

```{r warning=FALSE}
ggplotly(ggplot(all_data, aes(x = Gov.Corruption, y = Suicide_Rate_MF, Country = Country)) +
  geom_point(size = 2) +
  ylab("Suicide Rate (Male & Female)") +
  xlab("Government Corruption") +  
  theme_minimal())
```

### Generosity Score vs Suicide Rates

```{r warning=FALSE}
ggplotly(ggplot(all_data, aes(x = Generosity, y = Suicide_Rate_MF, Country = Country)) +
  geom_point(size = 2) +
  ylab("Suicide Rate (Male & Female)") +
  xlab("Generosity") +  
  theme_minimal())
```
